Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-06-05 paper/main.tex
cd7v
fixes theory section

\documentclass{uai2025} % for initial submission
%\documentclass[accepted]{uai2025} % after acceptance, for a revised version; 
% also before submission to see how the non-anonymous paper would look like

@@

We consider two models of a domain experts. Given two variables
$ X $ and $ Y $ that are part of an unknown causal DAG $G$, an \emph{strong ancestral oracle} 
$\mathcal{A}_G$ is defined as:
$$\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{None} & \textrm{otherwise} \\
\end{cases}$$

whereas a \emph{weak ancestral oracle} is defined as:
$$
\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{rand}(\{X \gets Y, X \to Y\}) & \textrm{otherwise} \\
\end{cases}
$$
where $\textrm{rand}(S)$ denotes a uniformly chosen element of $S$.
That is, our ancestral oracles only provide information on the direction of cause-effect
relationships, and do not distinguish between direct and indirect effects. 
The weak ancestral oracle is even unable to detect the absence of cause-effect
relationships between two variables. Using such oracles for DAG construction  
inherently risks that superfluous edges are created.

Our approach uses the data to decide where to add edges and which edges 
are superfluous and can be removed.
As a useful abstraction, let us assume that we
have access to a second oracle $\mathcal{D}_G$ than can answer d-separation
queries with respect to the unknown target graph:
$\mathcal{D}_G(X,Y,\mathbf{Z})=1$ iff $X$ and $Y$ are $d$-separated by $\mathbf{Z}$ in 
$G$. These are the standard oracles considered in constraint-based structure
learning algorithms, such as the PC algorithm \cite{Spirtes2001}.

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$B$,$k$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
	    $e \gets \mathcal{A}_G(X, Y)$ \;
	    \uIf{ $(V,E \cup L \cup e)$ is acyclic }{
            	$L \gets L \cup  \{ e \} $\;
	    } \Else {
            	$L \gets L \cup  \{ \textrm{reverse}(e) \} $\;
	    }
        }
	\If{$|L|\geq k$}{\KwGoTo 12\;}
    }
    add all edges in $L$ to $E$\;
    \Return{$(V,E)$}\;
}
\caption{Adding edges based on data and domain knowledge.}
\label{algo:expand}
\end{algorithm}

@@

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a weak ancestral oracle $\mathcal{A}_G$, 
$\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_G, \emptyset, \infty)
\supseteq G^+$.
\label{prop:weakexpand}
\end{proposition}

\begin{proof}
Every pair of vertices $X, Y$ that are $d$-connected 
in $G$ are now connected by an edge 
during $\textsc{Expand}$. For $X \in \textrm{An}_G(Y)$, the edge is 
$X \to Y$ like when using a strong ancestral oracle; 
since only correctly oriented
edges are added between ancestor-descendant pairs, 
such additions never create cycles.
For $X, Y$ that are only $d$-connected throug common ancestors, 
an arbitrary edge is added in a manner that preserves acyclicity.
The result is a supergraph of $G^+$.
\end{proof}

@@

\begin{theorem}
Given a d-separation oracle $\mathcal{D}_G$ and a strong or weak 
ancestral oracle $\mathcal{A}_G$, 
$\textsc{Prune}(\textsc{Expand}((V,\emptyset),\mathcal{D}_G,\mathcal{A}_G,\emptyset,\infty),\mathcal{D}_G)=G.$
\end{theorem}

>>>>

\documentclass[accepted]{uai2025} % for initial submission
%\documentclass[accepted]{uai2025} % after acceptance, for a revised version; 
% also before submission to see how the non-anonymous paper would look like

@@

We consider two models of a domain experts. Given two variables
$ X $ and $ Y $ that are part of a directed graph $G$, an \emph{strong ancestral oracle} 
$\mathcal{A}_G$ is defined as:
$$\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{None} & \textrm{otherwise} \\
\end{cases}$$
Note that the ancestral oracle does not consider differences between direct and 
indirect relationships: for any $G, H$ where $G^+=H^+$, we have $\mathcal{A}_G=\mathcal{A}_H$.

Experts can make mistakes. In our theoretical analysis, we will model two different types
of mistakes that the algorithm can recover from.

\begin{definition}
Let $G=(V,E_G)$ be a DAG let and $H=(V,E_H)$ be a directed graph (possibly cyclic) 
such that $E_G \subseteq E_H$. Then $\mathcal{A}_H$
is called a \emph{$G$-consistent ancestral oracle}. Further, 
if $H=(V,E_H)$ is a supergraph of $G$ that contains at least one cycle, then 
$\mathcal{A}_H$ is called a \emph{$G$-consistent ancestral oracle} with respect to $G$.
\end{definition}

These two oracles still provide correct information on all ancestral relationships that are 
present in $G$, but cannot reliably detect the \emph{absence} of ancestral relationships.
Further, the inconsistent ancestral oracle sometimes contradicts itself: there are at least two
variables $X,Y$ where its information implies both a causal effect of $X$ on $Y$ and a reverse 
causal effect of $Y$ on $X$. Therefore, if we use such oracles for DAG construction, we will
introduce superfluous edges into our graph.

Our approach uses the data to decide where to add edges and which edges 
are superfluous and can be removed.
As a useful abstraction, let us assume that we
have access to a second oracle $\mathcal{D}_G$ than can answer d-separation
queries with respect to the unknown target graph:
$\mathcal{D}_G(X,Y,\mathbf{Z})=1$ iff $X$ and $Y$ are $d$-separated by $\mathbf{Z}$ in 
$G$, and $0$ otherwise. These are the standard oracles considered in constraint-based structure
learning algorithms, such as the PC algorithm \cite{Spirtes2001}.

@@

\begin{algorithm}[h]
\DontPrintSemicolon
\SetAlgoLined
\SetKwFunction{Expand}{Expand}
\SetKw{KwGoTo}{go to}
\SetKwProg{Fn}{Function}{:}{}
\Fn{{\sc Expand}($V,E,\mathcal{D}_G,\mathcal{A}_G$,$B$,$k$)}{
    let $L \gets \{\}$\;
    \ForEach{$X, Y$ where $X \to Y \notin E$ and $Y \to X \notin E$ and 
	$\{X,Y\} \notin B$}{
        let $\mathbf{Z}$ be a set that $d$-separates $X$ and $Y$ in $(V,E)$\;
        \If{$\mathcal{D}_G(X, Y, \mathbf{Z}) = 0$}{
	    $e \gets \mathcal{A}_G(X, Y)$ \;
            $L \gets L \cup  \{ e \} $\;
        }
	\If{$|L|\geq k$}{\KwGoTo 13\;}
    }
    add all edges in $L$ to $E$\;
    \Return{$(V,E)$}\;
}
\caption{Adding edges based on data and domain knowledge.}
\label{algo:expand}
\end{algorithm}

@@

The following is immediately obtained as a corollary.

\begin{proposition}
For a conditional independence oracle
 $\mathcal{D}_G$ and a $G$-consistent ancestral oracle $\mathcal{A}_H$, 
$\textsc{Expand}(V,\emptyset,\mathcal{D}_G,\mathcal{A}_H, \emptyset, \infty)
=H^+ \supseteq G^+$.
\label{prop:weakexpand}
\end{proposition}

@@

\begin{theorem}
Given a d-separation oracle $\mathcal{D}_G$ and a $G$-consistent 
ancestral oracle $\mathcal{A}_H$, 
$\textsc{Prune}(\textsc{Expand}((V,\emptyset),\mathcal{D}_G,\mathcal{A}_H,\emptyset,\infty),\mathcal{D}_G)=G.$
\end{theorem}
